---
title: "scRNA File formats"
output: html_document
date: "2025-10-13"
---

Script to read and access single cell matrices in various file formats 
and converting to seurat object 

```{r}

setwd("../Fileformat/")

```

#Packages required
1. Seurat is the main R package for single-cell data analysis.
2. SeuratDisk is a helper package mainly for saving, loading, and converting Seurat objects efficiently across different formats and tools (like Python’s AnnData / Scanpy).

#If not installed
```{r}
install.packages("Seurat")
install.packages("remotes")
remotes::install_github("mojaveazure/seurat-disk")
```

#Load the libraries
```{r}
library(Seurat)
library(SeuratDisk)
```

FILE FORMATS OF scRNA

1. RDS format = Completely proceesed seurat object 
#Read using readRDS()
```{r include=FALSE}
rds_obj <- readRDS(file = "../Fileformat/training_session_datasets/ifnb_harmony.rds")
str(rds_obj)  #To see the structure of the object
```

```{r}
View(rds_obj) #To view the entire objects
View(rds_obj@meta.data) #To view the metadata
```

2. 10x cellranger - 

#2.1 HDF5 format = Raw Feature barcode matrix [ROws - features or genes, COLUMNS - Cell or barcodes ]

```{r}
#Step 1 - Read using Read10X_h5()
hdf5_raw_obj  <- Read10X_h5(filename = "../Fileformat/training_session_datasets/20k_NSCLC_DTC_3p_nextgem_Multiplex_count_raw_feature_bc_matrix.h5")
str(hdf5_raw_obj)
#It has 3 lists
```

```{r}
#Step 2 Extract only gene expression which contains feature barcode matrix
counts_hdf5 <- hdf5_raw_obj$`Gene Expression`
counts_hdf5[1:5,1:5]
```

```{r}
#Step 3- Create Seurat object
counts_hdf5_seurat <- CreateSeuratObject(counts = counts_hdf5, project = "NSCLC", min.cells = 3, min.features = 200)
```

# 2.2 HDF5 format = Filtered Feature barcode matrix [ROws - features or genes, COLUMNS - Cell or barcodes ]
```{r}
#Step 1 - Read using Read10X_h5()
hdf5_filtered_obj  <- Read10X_h5(filename = "../Fileformat/training_session_datasets/20k_PBMC_3p_HT_nextgem_Chromium_X_filtered_feature_bc_matrix.h5", 
                        use.names = TRUE, 
                        unique.features = TRUE)
#Step 2 - Create Seurat object
hdf5_filtered_seurat <- CreateSeuratObject(counts = hdf5_obj, project = "PBMC", min.cells = 3, 
                                  min.features = 200)

```

#2.3 MTX format - Proccesed files from cellranger have .mtx -  Matrix Market format file format - 
Extract 10k_PBMC_3p_nextgem_Chromium_Controller_raw_feature_bc_matrix.tar.gz 
Files are extracted in raw_feature_bc_matrix folder
```{r}
#Step 1 - Read by ReadMtx()
mtx_object <- ReadMtx(mtx = "../Fileformat/training_session_datasets/raw_feature_bc_matrix/matrix.mtx.gz", cells = "../Fileformat/training_session_datasets/raw_feature_bc_matrix/barcodes.tsv.gz", features = "../Fileformat/training_session_datasets/raw_feature_bc_matrix/features.tsv.gz")

#Step 2 - Create Seurat object
mtx_seurat <- CreateSeuratObject(counts = mtx_object)
```


3..loom files  - Used to hold large data of scRNA 

```{r}
#Step 1 - Open a loom file
loom_obj <- Connect(filename = "../Fileformat/training_session_datasets/adult-hem-organs-10X-bone-marrow.loom", mode = "r")

#Step 2 Convert to Seurat
loom_seurat <- as.Seurat(loom_obj)
```

4.Anndata  .h5ad format
```{r}
#Step 1 Convert h5ad to h5seurat
Convert("../Fileformat/adata_SS2_for_download.h5ad", dest = "h5seurat", overwrite = TRUE)
```
```{r}
#Step 2 - Load h5seuart a seurat obj
h5ad_seuratr <- LoadH5Seurat("../Fileformat/adata_SS2_for_download.h5seurat", images = FALSE)
```

